---
title: "List columns"
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
    comment = "#>"
)
pkgload::load_all()
library(pillar)
options(width = 160)
```

Parquet has a `LIST` logical type for columns whose values are variable-length
sequences of a scalar type. nanoparquet maps these to and from R list columns,
i.e. R list vectors whose elements are atomic vectors.

::: callout-note
Parquet supports nested lists (lists of lists), but nanoparquet cannot
currently read or write them. Only flat lists with scalar elements are
supported.
:::

## Supported element types

When *reading*, nanoparquet supports any scalar element type that it can
read in general (see [nanoparquet-types]), including `INT32`, `DOUBLE`,
`STRING`, `DATE`, `TIMESTAMP`, `BOOLEAN`, and so on. Nested `LIST`
elements and unsupported types such as `MAP` are not supported.

When *writing*, nanoparquet currently supports three element types:

| Parquet element type | R element type  |
|:---------------------|:----------------|
| `INT32`              | `integer`       |
| `DOUBLE`             | `double`        |
| `STRING`             | `character`     |

## Writing list columns

To write a data frame with a list column, put the column in the data
frame as a list. Each element of the list must be an atomic vector of a
supported type, `NULL` (missing value), or an empty vector. The elements
may contain missing values (`NA`).

```{r}
df <- data.frame(id = 1:4)
df$ints <- list(1L, c(2L, 3L), NULL, c(4L, NA_integer_, 6L))
df$doubles <- list(1.5, c(2.5, 3.5), NULL, c(4.5, NA_real_, 6.5))
df$strings <- list("a", c("b", "c"), NULL, c("d", NA_character_, "f"))
df
```

```{r}
tmp <- tempfile(fileext = ".parquet")
write_parquet(df, tmp)
```

## Reading list columns

`read_parquet()` reads `LIST` Parquet columns back as R list columns:

```{r}
as.data.frame(read_parquet(tmp))
```

The round-trip preserves both the structure and the missing value pattern.

## Missing values

nanoparquet distinguishes two kinds of missingness in list columns:

* `NULL` — the entire list entry is missing. In Parquet this is an
  absent (optional) outer list value.
* `NA` within an element vector — a missing scalar inside a present list.
  In Parquet this is an absent (optional) inner element value.
* An empty vector — the list is present but has zero elements.

```{r}
df2 <- data.frame(id = 1:4)
df2$x <- list(
    c(1L, 2L, 3L), # normal, three elements
    integer(0), # present, but empty
    NULL, # entirely missing
    c(4L, NA, 6L) # present, with an NA element
)
df2
```

```{r}
tmp2 <- tempfile(fileext = ".parquet")
write_parquet(df2, tmp2)
as.data.frame(read_parquet(tmp2))
```

## Inspecting the schema

`infer_parquet_schema()` shows how nanoparquet would encode each column:

```{r}
infer_parquet_schema(df)
```

A `LIST` column occupies three rows in the schema table: the outer list node
(logical type `LIST`), the middle repeated group node (`list`), and the leaf
element node (`element`). These three rows all share the same `r_col` value.

To inspect the schema of an existing file use `read_parquet_schema()`:

```{r}
read_parquet_schema(tmp)
```

The `r_type` column shows `list(integer)`, `list(double)`, and
`list(character)` for the three list columns.

## Explicit schema with `parquet_schema()`

nanoparquet infers the element type from the R list contents automatically.
You can also specify it explicitly with `parquet_schema()` using
`list("LIST", element = <type>)`:

```{r}
schema <- parquet_schema(
    id     = "INT32",
    values = list("LIST", element = "DOUBLE")
)
schema
```

```{r}
df3 <- data.frame(id = 1:3)
df3$values <- list(c(1.1, 2.2), c(3.3), NULL)
tmp3 <- tempfile(fileext = ".parquet")
write_parquet(df3, tmp3, schema = schema)
as.data.frame(read_parquet(tmp3))
```

## Dictionary encoding

List columns support dictionary encoding, which can reduce file size when
list elements repeat frequently:

```{r}
df4 <- data.frame(id = 1:6)
df4$x <- list(
    c(1L, 2L), c(2L, 3L), c(1L, 2L),
    c(2L, 3L), c(1L, 2L), c(2L, 3L)
)
tmp4 <- tempfile(fileext = ".parquet")
write_parquet(df4, tmp4, encoding = c(x = "RLE_DICTIONARY"))
as.data.frame(read_parquet(tmp4))
```

## Interoperability

nanoparquet can read standard `LIST` Parquet columns written by other
tools, such as Arrow or DuckDB. The standard three-layer representation
(`OPTIONAL` or `REQUIRED` list → `REPEATED` group → `OPTIONAL` or
`REQUIRED` element) is fully supported, for both data page version 1 and
version 2.

The four combinations of optional/required outer list and optional/required
element are all handled:

| Outer list   | Element      | `NULL` entry? | `NA` element? |
|:-------------|:-------------|:--------------|:--------------|
| `OPTIONAL`   | `OPTIONAL`   | yes           | yes           |
| `OPTIONAL`   | `REQUIRED`   | yes           | no            |
| `REQUIRED`   | `OPTIONAL`   | no            | yes           |
| `REQUIRED`   | `REQUIRED`   | no            | no            |

Empty vectors (e.g. `integer(0)`) are supported in all four combinations.

nanoparquet always writes `OPTIONAL`/`OPTIONAL` lists (i.e. both the list
entry and each element may be missing), which is the most general case and
is interoperable with all Parquet readers.
